---
title: "Matrices and Multivariate Normality"
author: "JDRS"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `car` library has some handy multivariate features as does the `heplots` library

```{r}
library(car)
library(heplots)

```
###  Examples of Matrices in R

There are plenty of built in matrix functions in R - R thinks in terms of matrices!

For starters, we create two 3x3 matrices and perform some basic function.

```{r}
#make the matrices
matrixA <- matrix(c(3, 2, 4, 7, 5, 0, 1, 0, 8), ncol = 3, byrow = T)
matrixB <- matrix(c(6, 1, 0, 2, 8, 7, 3, 4, 5), ncol = 3, byrow = T)

#Add a constant to a matrix
matrixA + 5

#Add two matrices
matrixA + matrixB

#Matrix multiplication
matrixA%*%matrixB

#transpose of a matrix
t(matrixA)

#inverse of a matrix
solve(matrixA)

#determinant of a matrix
det(matrixA)

#eigenvalues/eigenvectors of a matrix
eigen(matrixA)

matrixbB <- matrix(c(3, 1, 1, 2), nrow = 2, byrow = T)
eigen(matrixbB)

```

###  Assessing univariate normality in R - Normal Quantile Plots

There are several good functions for creating normal quantile plots.  Built in to R are `qqnorm` (and `qqline`).  I think better is `qqPlot()` in the `car()` package.

**Example**: World Bank Data

```{r}
#read in data from online location
WBtrans <- read.csv("http://reuningscherer.net/multivariate/data/WB.2016.WithTrans.csv", header = T)
names(WBtrans)

#Get a normal quantile plot of Energy Use per capita, raw scale
qqnorm(WBtrans$GNI)

#somewhat more refined plot
qqnorm(WBtrans$EnergyUse, col = 'red', pch = 19, cex = 1.2, main = "Normal Quantile Plot")
qqline(WBtrans$EnergyUse, lwd = 2)

#My preferred normal quantile plot function - qqPlot in the car package
qqPlot(WBtrans$EnergyUse, col = 'red', pch = 19)

#Same plot, log scale
qqPlot(WBtrans$logEnergyUse, col = 'red', pch = 19)

```

###  Assessing MULTIVARIATE normality in R - Chi-Square Quantile Plots

I have written a function called `CSQPlot()` that creates chi-square quantile plots from raw data.  Here's an example on some variables from the World Bank Data.

**Example**: World Bank Data

```{r}
#make square plots

#read in data from online location

#WBtrans <- read.csv("http://reuningscherer.net/multivariate/data/WB.2016.WithTrans.csv", header = T)
#names(WBtrans)

#check multivariate normality ??? get online function
source("http://www.reuningscherer.net/multivariate/R/CSQPlot.r.txt")

#Only works for countries with complete data
#Look at following variables : 
#	%Rural
#	log(CO2 per capita)
#	log(energy use per capita)
# log(GNI)
#	Fertility Rate
#	log(Imports)
#	log(Exports)

#  This leaves 119 countries with complete data
WBtemp <- WBtrans[complete.cases(WBtrans[, c(4, 30, 32, 31, 12, 33, 34)]), ]

#run the function on transformed data
CSQPlot(WBtemp[, c(4, 30, 32, 31, 12, 33, 34)], label = "Transformed 2016 World Bank Data")

#here is untransformed data
CSQPlot(WBtemp[, c(4, 18, 19, 5, 12, 7, 8)], label = "Raw 2016 World Bank Data")

```
Let's try doing same thing in two other packages: use `qqPlot()` in `car`.
```{r}
#First try in car package.  This uses the qqPlot function but requires that we first calculate the Mahalanobis distances (from the heplots 

#run the function on transformed data
qqPlot(Mahalanobis(WBtemp[, c(4, 30, 32, 31, 12, 33, 34)]), dist = "chisq", df = 7, main = "Transformed 2016 World Bank Data", pch=19, col = 'red', ylab = "Squared Mahalanobis Distances")


#Another option - use the cqplot function in 'heplots'.  Again, transformed data.  This function has options to change the confidence envelope, colors, etc.  Super cool function!

cqplot(WBtemp[, c(4, 30, 32, 31, 12, 33, 34)], main = "Transformed 2016 World Bank Data", pch = 19, conf = .95)
cqplot(WBtemp[, c(4, 30, 32, 31, 12, 33, 34)], main = "Transformed 2016 World Bank Data", pch = 19, conf = .95, fill.color = 'light blue', fill.alpha = .1)


```

The results for `cqplot()` and `CSQPlot()` are basically identical.  The envelope for the `qqPlot()` function after calculating squared Mahalanobis distances is a bit wider and is calculated slightly differently.



FINALLY - 

**SIMULATION EXAMPLE to see the effect of sample size on shape of Chi-Square Quantile plots**

```{r}
#an example of the effect of sample size on chi-square quantile plots 
#to test multivariate normality


#simulation to show effect of sample size
#p = number of dimensions
#requires function reschisqplot function

multnormsize <- function(p) {
  temp <- matrix(rnorm(p*100*p, 0, 1), nrow = 100*p, ncol = p)
  vec <- round(c(c(1:4), seq(0, 8, 1)*p+10, 20*p, 90*p), 0)+p
  for (i in 1:length(vec)){
    cqplot(temp[c(1:vec[i]),], main = paste("Dimensions = ",p," , N =", vec[i]))
  }
}

multnormsize(10)

```
```{r}
#try doing same thing in 
```

